---
title: "Authenticated requests"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Authenticated requests}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

```{r setup}
library(codeminer.api)
```

## When you need this

If the CodeMiner API server you are calling requires authentication
(production deployments typically do), you need to attach an identity to each
outgoing request. `codeminer.api` provides a small auth-strategy abstraction
so you can configure this once per session and then use the regular helpers
(`DESCRIPTION()`, `CODES()`, etc.) without thinking about auth on every call.

If the server you are calling does not require authentication (a local
development server with no `PLUMBER_SECRET`, for instance), nothing in this
vignette applies to you.

## Configuring auth (the common case)

There are three ways to configure auth, in roughly increasing order of
specificity. Pick whichever is most convenient for your workflow.

### 1. Environment variable (recommended for development)

Set `CODEMINER_DEV_USER` in your `.Renviron`:

```
CODEMINER_DEV_USER=alasdair
```

The package's default resolver picks this up and attaches an `X-Dev-User`
header to every outgoing request. No code changes needed:

```{r}
DESCRIPTION(pattern = "asthma")
```

This only works against servers running with the `CODEMINER_DEV_AUTH=1`
environment variable, which is intended for development setups. Production
servers ignore the `X-Dev-User` header.

### 2. Session-wide programmatic configuration

If you need to switch identities mid-session, or want a strategy that env
vars can't express, use `auth_use()`:

```{r}
auth_use(auth_dev_user("alasdair"))
DESCRIPTION(pattern = "asthma")  # X-Dev-User: alasdair attached
```

`auth_clear()` removes the session setting and falls back to the env-var
default.

### 3. Per-call override

Pass `auth = ...` directly to any helper. This wins over both the env var
and the session setting:

```{r}
DESCRIPTION(pattern = "asthma", auth = auth_dev_user("admin"))
```

## Checking what's configured

When something returns 401, the first question is "what auth am I actually
sending?". Two introspection helpers answer this:

```{r}
auth_current()
#> <codeminer_auth>
#>   kind: dev_user
#>   Dev user 'alasdair' (X-Dev-User header)

auth_describe()
#> [1] "Dev user 'alasdair' (X-Dev-User header)"
```

`auth_current()` returns the strategy that the next helper call would use
(option from `auth_use()` if set, otherwise the env-var-driven default).
`auth_describe()` returns just the description string, useful for embedding
in your own logs or error messages.

## Custom auth strategies

If your server uses an auth scheme that the package doesn't ship a
constructor for (institutional SSO, an API key, a bearer token from a
separate auth service), build your own with `auth_custom()`.

A strategy is just a function that takes an `httr2::request` and returns a
(possibly modified) `httr2::request` — the same shape as `httr2::req_auth_*`.
Wrap it with `auth_custom()` to get the package's printing and introspection
support, then pass it to `auth_use()`:

```{r}
my_auth <- auth_custom(
  function(req) httr2::req_auth_bearer_token(req, Sys.getenv("MY_TOKEN")),
  description = "Bearer token from MY_TOKEN env var"
)
auth_use(my_auth)
```

### Worked example: token from a separate auth service

Some institutional setups require the client to exchange a long-lived
credential (username + API key, say) for a short-lived bearer token by
calling a separate auth endpoint, then send that token with each request.

The strategy pattern handles this naturally because closures can hold cached
state across calls — you don't need to re-fetch the token for every request:

```{r}
my_org_token_service_auth <- function(
  username = Sys.getenv("MYORG_USERNAME"),
  api_key = Sys.getenv("MYORG_API_KEY"),
  token_url = "https://auth.myorg.internal/token"
) {
  force(username)
  force(api_key)
  force(token_url)

  cached_token <- NULL
  cached_until <- 0

  auth_custom(
    function(req) {
      now <- unclass(Sys.time())

      # Refresh proactively, before the cached token expires
      if (is.null(cached_token) || now > cached_until - 30) {
        resp <- httr2::request(token_url) |>
          httr2::req_body_json(list(
            username = username,
            api_key = api_key
          )) |>
          httr2::req_perform() |>
          httr2::resp_body_json()
        cached_token <<- resp$access_token
        cached_until <<- now + resp$expires_in
      }

      httr2::req_auth_bearer_token(req, cached_token)
    },
    description = paste0("Bearer token from ", token_url, " (cached)")
  )
}

auth_use(my_org_token_service_auth())
```

Two patterns worth noting:

- **`<<-` updates the variables in the enclosing scope** so the cache
  persists across calls. Without `<<-`, each call would create new local
  variables and the cache would never warm up.
- **`force()` captures parameters eagerly.** R's lazy evaluation would
  otherwise capture a *promise* to evaluate the parameter; if the user
  later reassigned the variable they passed in, the closure would see the
  new value. `force()` is the standard guard.

## The strategy contract

For anyone writing their own strategy, the rules:

> A strategy receives an `httr2::request` immediately before it is performed,
> with URL and body already set. It may modify any aspect — headers, options,
> even the URL — and must return the modified request. Strategies should not
> perform the request themselves.

The order of resolution at call time is:

1. The `auth = ...` argument passed directly to the helper, if given.
2. Otherwise, the strategy set by `auth_use()`, if any.
3. Otherwise, `default_auth()` — which returns `auth_dev_user()` if
   `CODEMINER_DEV_USER` is set, or `auth_none()` otherwise.
