---
title: "Introduction to codeminer.api"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to codeminer.api}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Overview

`codeminer.api` provides two things:

1. A **REST API server** that exposes all `codeminer` query functions as HTTP
   endpoints.
2. **Client wrapper functions** (`CODES()`, `DESCRIPTION()`, `MAP()`, etc.)
   that are drop-in replacements for the equivalent `codeminer` functions --
   same signatures, same outputs, same messages and errors -- except the work
   happens on the server.

## Setup

### 1. Set the database path

Point `CODEMINER_DB_PATH` at your CodeMiner database:

```{r setup-db}
# Set for the current session
Sys.setenv(CODEMINER_DB_PATH = "/path/to/your/codeminer.duckdb")

# Or create a dummy database for testing
db_path <- tempfile(fileext = ".duckdb")
codeminer::create_dummy_database(db_path)
Sys.setenv(CODEMINER_DB_PATH = db_path)
```

### 2. Start the API server

```{r start-server}
library(codeminer.api)

# Start in background (returns immediately)
bg <- run_codeminer_api(
  host = "127.0.0.1",
  port = 8000,
  background = TRUE
)

# Check it's alive
bg$is_alive()
```

### 3. Point the client at the server

```{r set-url}
options(codeminer.api.url = "http://127.0.0.1:8000")
```

## Using the client wrappers

The client wrappers mirror `codeminer` exactly. If you already know `codeminer`,
you know `codeminer.api`:

### Search descriptions

```{r description}
# Find ICD-10 codes mentioning "asthma"
DESCRIPTION("asthma", type = "ICD-10")
```

### Look up specific codes

```{r codes}
# Single code
CODES("J45", type = "ICD-10")

# Multiple codes
CODES("J45", "E11", "I10", type = "ICD-10")

# Pipe-separated syntax works too
CODES("J45 || E11 || I10", type = "ICD-10")
```

### Search codes by pattern

```{r codes-like}
# Regex match on code values
CODES_LIKE("^J45", type = "ICD-10")
```

### Hierarchy traversal

```{r hierarchy}
# All descendants (transitive children)
CHILDREN("J45", type = "ICD-10")

# Immediate children only
N_CHILDREN("J45", depth = 1, type = "ICD-10")

# All ancestors
PARENTS("J450", type = "ICD-10")

# Immediate parents only
N_PARENTS("J450", depth = 1, type = "ICD-10")
```

### Attributes

```{r attributes}
# Get attributes for codes
ATTRIBUTES_FOR("73211009", type = "sct")

# Get codes that have certain attributes
HAS_ATTRIBUTES("73211009", type = "sct")
```

### Relationship types

```{r relationship-types}
# What relationship types originate from these codes?
RELATIONSHIP_TYPES_FROM("73211009", type = "sct")

# What relationship types point to these codes?
RELATIONSHIP_TYPES_TO("73211009", type = "sct")
```

### Code mapping

```{r map}
# Map ICD-10 codes to Read 2
MAP("J45", from = "ICD-10", to = "Read 2")
```

### Database metadata

```{r metadata}
# Discover what code types and versions are available
get_codeminer_metadata()

# Just lookup tables
get_codeminer_metadata("lookup")

# Lookup and mapping tables
get_codeminer_metadata(c("lookup", "mapping"))
```

## Transparency: API vs local

The client wrappers are designed to be transparent. You should not be able to
tell the difference between calling through the API and calling `codeminer`
directly:

```{r transparency}
# These should produce identical results
codeminer.api::CODES("J45", type = "ICD-10")
codeminer::CODES("J45", type = "ICD-10")

# Errors and warnings are also reproduced faithfully
codeminer.api::CODES("nonexistent", type = "ICD-10")
codeminer::CODES("nonexistent", type = "ICD-10")
```

## Using the REST API directly

You can also call the endpoints directly with `httr2` or `curl`. All data
endpoints accept POST requests with a JSON body:

```{r httr2-direct}
library(httr2)

response <- request("http://127.0.0.1:8000/DESCRIPTION") |>
  req_method("POST") |>
  req_body_json(list(pattern = "asthma", type = "ICD-10")) |>
  req_perform()

resp_body_json(response, simplifyVector = TRUE)
```

Or from the command line:

```bash
curl -X POST http://127.0.0.1:8000/DESCRIPTION \
  -H "Content-Type: application/json" \
  -d '{"pattern": "asthma", "type": "ICD-10"}'
```

### Available endpoints

| Endpoint | Parameters |
|----------|-----------|
| `GET /health` | (none) |
| `POST /DESCRIPTION` | `pattern`, `type`, `lookup_version`, `ignore_case`, `preferred_description_only`, `col_filters` |
| `POST /CODES` | `codes`, `type`, `lookup_version`, `preferred_description_only`, `col_filters` |
| `POST /CODES_LIKE` | `pattern`, `type`, `lookup_version`, `preferred_description_only`, `col_filters` |
| `POST /CHILDREN` | `codes`, `type`, `lookup_version`, `relationship_version`, `preferred_description_only`, `col_filters` |
| `POST /N_CHILDREN` | `codes`, `depth`, `type`, `lookup_version`, `relationship_version`, `preferred_description_only`, `col_filters` |
| `POST /PARENTS` | `codes`, `type`, `lookup_version`, `relationship_version`, `preferred_description_only`, `col_filters` |
| `POST /N_PARENTS` | `codes`, `depth`, `type`, `lookup_version`, `relationship_version`, `preferred_description_only`, `col_filters` |
| `POST /ATTRIBUTES_FOR` | `codes`, `type`, `lookup_version`, `relationship_version`, `relationship_types`, `preferred_description_only`, `col_filters` |
| `POST /HAS_ATTRIBUTES` | `codes`, `type`, `lookup_version`, `relationship_version`, `relationship_types`, `preferred_description_only`, `col_filters` |
| `POST /RELATIONSHIP_TYPES_FROM` | `codes`, `type`, `relationship_version`, `col_filters` |
| `POST /RELATIONSHIP_TYPES_TO` | `codes`, `type`, `relationship_version`, `col_filters` |
| `POST /MAP` | `codes`, `from`, `to`, `map_version`, `lookup_version`, `col_filters` |
| `GET /metadata` | `type` (optional, comma-separated: `lookup`, `mapping`, `relationship`) |

### Swagger UI

When the server is running with `docs = TRUE` (the default), interactive API
documentation is available at `http://127.0.0.1:8000/__docs__/`.

## Checking the connection

```{r check-connection}
# Diagnostic function
check_api_connection()
```

## Stopping the server

```{r stop}
bg$kill()
bg$is_alive()
```

## Docker deployment

For production use, the package includes a `Dockerfile` and
`docker-compose.yml`:

```bash
# Build and run
docker-compose up -d

# Mount your real database
# Edit docker-compose.yml to map your data directory:
#   volumes:
#     - /path/to/your/data:/data
```

## Troubleshooting

**"CODEMINER_DB_PATH not set"** -- Set the environment variable before starting
the server:

```{r troubleshoot-db}
Sys.setenv(CODEMINER_DB_PATH = "/path/to/your/codeminer.duckdb")
```

**"Failed to connect to CodeMiner API"** -- The server isn't running or the URL
is wrong:

```{r troubleshoot-connect}
# Check the configured URL
getOption("codeminer.api.url")

# Test the connection
check_api_connection()
```

**Background process died** -- Check the error output:

```{r troubleshoot-bg}
if (!bg$is_alive()) {
  cat(bg$read_all_error())
}
```
