---
title: "Introduction to codeminer.api"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to codeminer.api}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

`codeminer.api` provides a REST API wrapper for the `codeminer` package, allowing you to access clinical code mappings through HTTP endpoints. This vignette demonstrates how to set up and use the API server.

## Setting Up a Database

For this tutorial, we'll use a dummy database. In production, you would point to your actual CodeMiner database.

```{r setup}
library(codeminer.api)

# Create a temporary dummy database
db_path <- tempfile(fileext = ".duckdb")
codeminer::create_dummy_database(db_path)

# Set the environment variable
Sys.setenv(CODEMINER_DB_PATH = db_path)
```

## Running the API Server

### Foreground Mode

Start the server in foreground mode for interactive use. This will block your R session until you stop it with Ctrl+C or the Stop button in RStudio.

```{r foreground, eval = FALSE}
# Start on default port 8000
run_codeminer_api()

# Custom configuration
run_codeminer_api(
  host = "127.0.0.1",
  port = 9000,
  docs = TRUE
)
```

### Background Mode

For most use cases, you'll want to run the API in the background:

```{r background, eval = FALSE}
# Start in background (do not run during vignette build)
bg <- run_codeminer_api(
  background = TRUE,
  port = 8888,
  quiet = TRUE
)

# Check if server is running
bg$is_alive()

# Get process ID
bg$get_pid()
```

## Using the API Endpoints

Once the server is running, you can make HTTP requests to access the data.

### Search by Description Pattern

The `/DESCRIPTION` endpoint searches for codes matching a text pattern:

```{r description_endpoint, eval = FALSE}
library(httr2)

response <- request("http://127.0.0.1:8888/DESCRIPTION") |>
  req_url_query(
    pattern = "asthma",
    code_type = "icd10"
  ) |>
  req_perform()

# Parse JSON response
results <- resp_body_json(response)
str(results)
```

### Lookup Specific Codes

The `/CODES` endpoint retrieves descriptions for specific codes:

```{r codes_endpoint, eval = FALSE}
# Example: query the CODES endpoint (do not run during vignette build)
response <- request("http://127.0.0.1:8888/CODES") |>
  req_url_query(
    codes = "J45",
    code_type = "icd10"
  ) |>
  req_perform()

results <- resp_body_json(response)
str(results)
```

You can also query multiple codes at once by providing a comma-separated string:

```{r codes_multiple, eval = FALSE}
# Example: query multiple codes (do not run during vignette build)
response <- request("http://127.0.0.1:8888/CODES") |>
  req_url_query(
    codes = c("J45", "E11", "I10"),
    code_type = "icd10",
    .multi = "explode"
  ) |>
  req_perform()

results <- resp_body_json(response)
str(results)
```

## Managing the Background Process

### Checking Status

```{r status, eval = FALSE}
# Check process status (do not run during vignette build)
# Is the process still running?
bg$is_alive()

# Get stdout/stderr
if (bg$is_alive()) {
  # Read recent output
  cat(bg$read_output())
  
  # Read recent errors (if any)
  cat(bg$read_error())
}
```

### Stopping the Server

```{r stop, eval = FALSE}
# Stop the server (do not run during vignette build)
bg$kill()

# Verify it stopped
bg$is_alive()
```

## Interactive API Documentation

When the server runs with `docs = TRUE` (the default), you can access Swagger UI for interactive API testing:

1. Start the server: `run_codeminer_api()`
2. Open your browser to: `http://127.0.0.1:8000/__docs__/`
3. Try out endpoints directly in the browser

To disable documentation (e.g., in production):

```{r no_docs, eval = FALSE}
run_codeminer_api(docs = FALSE)
```

## Complete Workflow Example

Here's a complete example workflow:

```{r workflow, eval = FALSE}
# Complete workflow example (do not run during vignette build)
# 1. Set up database
temp_db <- tempfile(fileext = ".duckdb")
codeminer::create_dummy_database(temp_db)
Sys.setenv(CODEMINER_DB_PATH = temp_db)

# 2. Start API server
api_process <- run_codeminer_api(
  background = TRUE,
  port = 8889,
  quiet = TRUE
)

# 3. Query the API
library(httr2)

# Search for diabetes codes
diabetes_response <- request("http://127.0.0.1:8889/DESCRIPTION") |>
  req_url_query(pattern = "diabetes", code_type = "icd10") |>
  req_perform()

diabetes_codes <- resp_body_json(diabetes_response)
print(diabetes_codes)

# Lookup specific codes
lookup_response <- request("http://127.0.0.1:8889/CODES") |>
  req_url_query(
    codes = c("J45", "E11", "I10"),
    code_type = "icd10",
    .multi = "explode"
  ) |>
  req_perform()

code_info <- resp_body_json(lookup_response)
str(code_info)

# 4. Clean up
api_process$kill()
api_process$is_alive() # verify it stopped
file.remove(temp_db)
```

## Troubleshooting

### Database Path Not Set

If you see errors about missing database paths:

```{r troubleshoot_db, eval = FALSE}
# Check if set
Sys.getenv("CODEMINER_DB_PATH")

# Set it
Sys.setenv(CODEMINER_DB_PATH = "/path/to/database.duckdb")
```

### Background Process Failed to Start

Check the error output:

```{r troubleshoot_bg, eval = FALSE}
if (!bg$is_alive()) {
  cat("Error output:\n", bg$read_error(), "\n")
  cat("Standard output:\n", bg$read_output(), "\n")
}
```

```{r cleanup, include = FALSE}
# Clean up demo database
if (file.exists(db_path)) {
  file.remove(db_path)
}
```
