---
title: "Developer Guide"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Developer Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Overview

The `codeminer.api` package has two primary responsibilities:

1. To expose functions from the `codeminer` package as REST API endpoints using `plumber`.
2. To provide client wrapper functions (`CODES()`, `DESCRIPTION()`, etc.) that behave identically to the original `codeminer` functions — including their outputs, user-facing messages, warnings, and errors.

The client wrappers are designed to be drop-in replacements for the local `codeminer` functions, except that computation is performed by the API.

# API Architecture

The API router is constructed by `create_codeminer_api()`, which sequentially adds endpoint helper functions. Each endpoint:

- is defined in `R/endpoints.R`
- wraps a backend `codeminer::*` function
- uses a unified handler factory (`codeminer_handler_factory()`) to ensure consistent condition-capture, error formatting, and JSON structure.

The overall structure:

```
create_codeminer_api()
 ├── add_health_endpoint()
 ├── add_description_endpoint()
 ├── add_codes_endpoint()
```

# Adding New Endpoints

To add a new endpoint:

## 1. Create endpoint helper

Add a new function to `R/endpoints.R`:

```
add_myendpoint_endpoint <- function(pr) {
  pr$handle(
    "GET",
    "/myendpoint",
    handler = codeminer_handler_factory(
      function(arg1, arg2) {
        codeminer::my_function(arg1, arg2)
      }
    )
  )
  pr
}
```

## 2. Register the endpoint

Add your helper to `create_codeminer_api()`:

```
create_codeminer_api <- function() {
  pr <- plumber::pr()

  pr <- add_health_endpoint(pr)
  pr <- add_description_endpoint(pr)
  pr <- add_codes_endpoint(pr)
  
  # new endpoint
  pr <- add_myendpoint_endpoint(pr)

  pr
}
```

## 3. Add a unit test

Use the standard helper:

```
test_endpoint_addition(
  add_myendpoint_endpoint,
  "myendpoint",
  "/myendpoint"
)
```

This checks that:

- the endpoint function exists and can be called
- the path is added to the router
- the router exposes the expected route

# Condition Handling and Design Rationale

`codeminer` backend functions use custom wrappers around `cli_inform()`, `cli_warn()`, and `cli_abort()` to create structured conditions with custom classes such as:

- `codeminer_message`
- `codeminer_warning`
- `codeminer_error`

These wrappers embed:

- original CLI message vectors (e.g., `c("i" = "info", "!" = "warn")`)
- the primary class for dispatch
- optional subclasses for more specific error types

The original message vector is stored in a `cli_message` attribute.

The API must:

1. Capture these structured conditions reliably;
2. Serialise them to JSON in a deterministic way;
3. Allow the client wrapper to reconstruct the original CLI structure exactly.

# How Backend Conditions Are Captured

All API endpoint calls are wrapped using `codeminer_handler_factory()`, which delegates to `codeminer_handle()`.

`codeminer_handle()`:

- runs the backend expression under `withCallingHandlers()`;
- intercepts:
  - `codeminer_warning`
  - `codeminer_message`
- extracts the message text or the `cli_message` attribute via `capture_cm_condition()`;
- stores conditions in a nested list structure;

Example pipeline:

```r
codeminer_handle(
  withCallingHandlers(
    backend_expression,
    codeminer_warning = capture_cm_condition("warnings", warn_env),
    codeminer_message = capture_cm_condition("messages", warn_env)
  ),
  res
)
```

Warnings and messages are captured into:

```r
warnings = list(
  "Warnings: N",
  list("!", "text A"),
  list("i", "text B"),
  ...
)
```


A similar structure is built for `messages`.

Errors are formatted by S3 method dispatch via `format_backend_error()`:

- default R errors -> 500 status
- codeminer errors -> 422 status, preserving `cli_message`

# JSON Output Schema

Every successful API response returns:

```json
{
  "result": <data structure returned by codeminer>,
  "warnings": [ list entries ],
  "messages": [ list entries ]
}
```


Every error response returns:

```
{
  "error": {
    "error_type": "<class>",
    "error_message": { named cli message vector }
},
  "warnings": [...],
  "messages": [...]
}
```

This makes the schema consistent regardless of success, warning, or error.

# Client-Side Reconstruction of Messages and Errors

Client wrapper functions:

1. Make API requests using `api_request()`;
2. Parse JSON into R structures using `httr2::resp_body_json()`;
3. Reconstruct CLI messages and warnings via:

```{r eval=FALSE}
print_captured_warnings_and_messages(parsed)
```

4. Print them using `cli_inform()` / `cli_warn()` to ensure identical output.

Errors are replayed using:

```{r eval=FALSE}
cli::cli_abort(
  convert_captured_message_to_cli_message_vector(error_list),
  class = parsed$error$error_type[[1]],
  call = call
)
```

Thus, from the R user’s perspective, calling:

```r
codeminer.api::CODES(...)
```

feels identical to using the local:

```r
codeminer::CODES(...)
```


# Background Mode During Development

Running the API in background mode (`run_codeminer_api(background = TRUE)`) launches the server in a fresh R session.

Important implications:

- Any local changes to the package source require reinstalling before restarting the background API.
- Environment variables (especially `CODEMINER_DB_PATH`) are inherited from the parent process.

# TODO / Future Enhancements

- **Switch all endpoints from GET -> POST**  
  Allows larger payloads (e.g. long code lists), avoids encoding issues with regex in query strings.
